---
title: "rbesd workflow: data preparation, quality checks, regression, and MrP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rbesd workflow: data preparation, quality checks, regression, and MrP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
set.seed(1)
```

This vignette walks through a realistic **rbesd** workflow using the bundled `data_demo`
dataset (3,000 respondents across five countries: Brazil, Mexico, Senegal, Philippines,
The UK):

1. [Dictionaries](#1-dictionaries) — how item and demographic dictionaries control harmonisation
2. [Data preparation](#2-data-preparation) — `as_besd()`, mapping, missing tokens, multichoice specs
3. [Data quality](#3-data-quality) — missing summaries, rare levels, recoding
4. [Summaries and plots](#4-summaries-and-plots) — response distributions, demographics, top-box, spider
5. [Regression](#5-regression) — by-country, multilevel, ordinal, multichoice, Bayesian
6. [Poststratification (MrP)](#6-poststratification-mrp) — population-weighted estimates

Bayesian and MrP sections are shown as code but not evaluated by default (they require
`brms` and a Stan toolchain). Set the environment variable `RBESD_RUN_BAYES_VIGNETTE=true`
to evaluate them.

```{r}
library(rbesd)
library(dplyr)
```

---

## 1. Dictionaries

### BeSD item dictionary

The BeSD dictionary describes every standard item: its ID, type, allowed response levels,
whether it is reverse-coded for top-box scoring, and the question text.

```{r}
besd_dict <- besd_dictionary()

besd_dict %>%
  select(item_id, domain, item_type, reverse) %>%
  print(n = 15)
```

Inspect a specific item's allowed levels:

```{r}
besd_dict %>%
  filter(item_id %in% c("tf_benefits", "sp_peer", "pr_reasons_ease_access")) %>%
  select(item_id, item_type, levels)
```

### Demographic dictionary

`dem_dictionary()` covers age and gender. Custom variables — such as ethnicity — can be
added with `modify_dictionary()`.

```{r}
dem_dictionary()
```

---

## 2. Data preparation

### Load the demo dataset

`data_demo` is a plain data frame with columns already named using BeSD item IDs. It ships
with the package and is used throughout this vignette.

```{r}
data("data_demo", package = "rbesd")
dplyr::glimpse(data_demo[, 1:10])
```

### Extend the demographic dictionary

`data_demo` contains an `dem_eth` (ethnicity) column that is not in the default
`dem_dictionary()`. Register it before calling `as_besd()`:

```{r}
dem_eth_df <- tibble::tibble(
  item_id        = "dem_eth",
  domain         = "predictor",
  item_type      = "categorical",
  levels         = list(sort(unique(data_demo$dem_eth))),
  reverse        = NA,
  question       = "What is your ethnicity?",
  question_short = "Ethnicity"
)

dem_dict <- modify_dictionary(
  df      = dem_eth_df,
  dict    = dem_dictionary(),
  replace = FALSE
)

dem_dict
```

### Create the `besd_data` object

When your columns are already named with standard BeSD item IDs (as in `data_demo`), no
`mapping` argument is needed. Declare free-text missing tokens via `missing_tokens` so they
are recoded to `NA`:

```{r}
y <- as_besd(
  df             = data_demo,
  country_col    = "country",
  dem_dict       = dem_dict,
  missing_tokens = list(.all = c("Prefer not to say", "Don't Know")),
  missing_action = "na"
)

print(y)
```

#### Using a mapping for raw exports

If your raw data uses survey-specific column names, supply a named character vector mapping
raw names to BeSD item IDs:

```r
mapping <- c(
  B1         = "tf_benefits",
  B2         = "tf_safety",
  B5         = "sp_peer",
  S3_RECODE  = "dem_age",
  S2         = "dem_gen",
  # multichoice: multiple source columns → one item ID
  setNames(rep("pr_reasons_ease_access", 12), paste0("A_B16_", 1:12))
)

y <- as_besd(
  df             = raw_data,
  country_col    = "country",
  dem_dict       = dem_dict,
  mapping        = mapping,
  missing_tokens = list(.all = c("Prefer not to say", "Don't Know")),
  missing_action = "na"
)
```

#### Multichoice items with text-prefix encoding

For multichoice items stored as packed text strings with a "Not " prefix indicating
non-selection, use `multichoice_specs`:

```r
y <- as_besd(
  ...,
  multichoice_specs = list(
    pr_reasons_ease_access = list(encoding = "text_prefix", not_prefix = "Not "),
    pr_service_quality     = list(encoding = "text_prefix", not_prefix = "Not ")
  )
)
```

### Validate the object

```{r}
besd_validate(y)
```

---

## 3. Data quality

### Missing data summary

```{r}
besd_missing_summary(y)
```

The joint listwise fraction shows how many rows complete-case deletion will remove across
all BeSD items.

### Response distributions and demographics

```{r}
smry <- summary(y)
print(smry, n = 15)
```

```{r}
besd_summary_demographics(smry)
```

### Check for rare factor levels

Thin cells within a country can cause near-separation in regression models. Inspect before
fitting — ethnicity is typically the most problematic:

```{r}
rl <- besd_rare_levels(y, predictors = c("dem_eth", "dem_gen", "dem_age"))
print(rl %>% arrange(country), n = Inf)
```

### Recode rare or unwanted levels

Set a level to `NA` and drop it from the factor:

```{r}
y <- besd_recode_missing(
  y,
  tokens      = list(dem_gen = "Other"),
  drop_levels = TRUE
)
```

Merge sparse ethnicity groups into "Other" to avoid near-separation:

```{r}
y <- besd_recode_levels(
  y,
  col    = "dem_eth",
  recode = c(
    # Brazil
    "Oriental"   = "Other",
    "Indigenous" = "Other",
    # Mexico
    "Afro-descent / black" = "Other",
    "Mulatto"              = "Other",
    "Romani / Gypsy"       = "Other",
    # Philippines
    "Tausog"       = "Other",
    "Maguindanao"  = "Other"
  )
)
```

Re-inspect to confirm:

```{r}
rl2 <- besd_rare_levels(y, predictors = c("dem_eth", "dem_gen", "dem_age"))
print(rl2 %>% arrange(country), n = Inf)
```

---

## 4. Summaries and plots

### Response distributions

```{r, fig.width=7, fig.height=5}
smry <- summary(y)

plot_besd_bars(
  smry %>% filter(item_id %in% c("tf_benefits", "tf_safety", "sp_peer"))
)
```

### Top-box summary

```{r}
tb <- summary(y, combine_top = TRUE)
tb %>%
  select(country, item_id, response, pct) %>%
  print(n = 15)
```

### Spider / radar plot

```{r, fig.width=7, fig.height=6}
plot_besd_spider(
  smry,
  item_ids = c("tf_benefits", "tf_safety", "sp_peer", "sp_family",
               "pr_recall", "pr_know_where"),
  compare  = "all"
)
```

### Ranked items

```{r, fig.width=7, fig.height=5}
plot_besd_ranked(smry)
```

---

## 5. Regression

All regression is handled by `besd_regress()`. Choose `scope` (`"by_country"` or
`"multilevel"`) and `engine` (`"frequentist"` or `"bayes"`). Extract results with
`tidy_model()`.

### 5.1 By-country, frequentist (binary outcome)

Fits a separate logistic regression per country:

```{r}
fit_bc <- besd_regress(
  y,
  outcome    = "sp_peer",
  predictors = c("dem_gen", "dem_age"),
  scope      = "by_country",
  engine     = "frequentist"
)

# Tidy fixed effects
tidy_model(fit_bc) %>% print(n = 15)
```

Exponentiate to get odds ratios:

```{r}
tidy_model(fit_bc, exponentiate = TRUE) %>% print(n = 15)
```

### 5.2 Multilevel, frequentist (binary outcome)

A single model with country as the random-effect grouping variable. `predictors` can split
into `common` (pooled across countries) and `context` (country-specific dummies, dropped
when the within-country cell count falls below `min_n_context`):

```{r}
fit_ml <- besd_regress(
  y,
  outcome       = "sp_peer",
  predictors    = list(common = c("dem_gen", "dem_age"), context = "dem_eth"),
  scope         = "multilevel",
  engine        = "frequentist",
  min_n_context = 10L
)

# Fixed effects + random SD parameters + country BLUPs
tidy_model(fit_ml, include_random = TRUE) %>% print(n = 20)
```

### 5.3 Ordinal outcome

`besd_regress()` detects the item type from the dictionary and switches to a cumulative
logit model automatically:

```{r}
fit_ord <- besd_regress(
  y,
  outcome    = "tf_benefits",
  predictors = c("dem_gen", "dem_age"),
  scope      = "by_country",
  engine     = "frequentist"
)

tidy_model(fit_ord) %>% print(n = 15)
```

### 5.4 Multichoice outcome

Multichoice items are split into one binary column per response option. `besd_regress()`
fits a separate logistic model for each option automatically. The `level` column in the
tidy output contains the response label:

```{r}
fit_mc <- besd_regress(
  y,
  outcome    = "pr_reasons_ease_access",
  predictors = c("dem_gen", "dem_age"),
  scope      = "by_country",
  engine     = "frequentist"
)

tidy_model(fit_mc) %>% print(n = 20)
```

### 5.5 Looping over multiple outcomes

```{r}
outcomes <- c("sp_peer", "sp_family", "pr_recall")

fits <- lapply(outcomes, function(yy) {
  besd_regress(
    y,
    outcome    = yy,
    predictors = list(common = c("dem_gen", "dem_age"), context = "dem_eth"),
    scope      = "multilevel",
    engine     = "frequentist"
  )
})
names(fits) <- outcomes

results <- dplyr::bind_rows(lapply(fits, tidy_model))
print(results, n = 20)
```

### 5.6 Bayesian multilevel

Requires `brms` and a configured Stan toolchain. First-time compilation takes ~1–2 minutes.
Set `RBESD_RUN_BAYES_VIGNETTE=true` to evaluate.

```{r, eval = requireNamespace("brms", quietly = TRUE) && Sys.getenv("RBESD_RUN_BAYES_VIGNETTE") == "true"}
fit_bayes <- besd_regress(
  y,
  outcome       = "sp_peer",
  predictors    = list(common = c("dem_gen", "dem_age"), context = "dem_eth"),
  scope         = "multilevel",
  engine        = "bayes",
  chains        = 2,
  iter          = 2000,
  cores         = 2,
  control       = list(adapt_delta = 0.95)
)

td <- tidy_model(fit_bayes)

# Check convergence: Rhat < 1.01 and ESS > 400 indicate reliable inference
td[, c("variable", "level", "country", "rhat", "ess")]
```

**Random slopes** allow the effect of common predictors to vary by country.
`correlated_re = FALSE` fits independent variances — more stable with few countries:

```{r, eval = requireNamespace("brms", quietly = TRUE) && Sys.getenv("RBESD_RUN_BAYES_VIGNETTE") == "true"}
fit_rs <- besd_regress(
  y,
  outcome        = "sp_peer",
  predictors     = list(common = c("dem_gen", "dem_age"), context = "dem_eth"),
  scope          = "multilevel",
  engine         = "bayes",
  random_slopes  = TRUE,
  correlated_re  = FALSE,
  chains         = 2,
  iter           = 2000,
  cores          = 2,
  control        = list(adapt_delta = 0.99)
)

td_rs <- tidy_model(fit_rs, include_random = TRUE)

# SD parameters (one per random-effect term)
td_rs[td_rs$param_type == "sd",
      c("variable", "level", "estimate", "lower", "upper")]

# Country-level BLUPs
td_rs[td_rs$param_type == "ranef",
      c("variable", "level", "country", "estimate", "lower", "upper")]
```

For large models, the `cmdstanr` backend is faster and supports within-chain parallelism:

```{r, eval = requireNamespace("brms", quietly = TRUE) && requireNamespace("cmdstanr", quietly = TRUE) && Sys.getenv("RBESD_RUN_BAYES_VIGNETTE") == "true"}
fit_cmd <- besd_regress(
  y,
  outcome        = "sp_peer",
  predictors     = list(common = c("dem_gen", "dem_age"), context = "dem_eth"),
  scope          = "multilevel",
  engine         = "bayes",
  random_slopes  = TRUE,
  correlated_re  = FALSE,
  backend        = "cmdstanr",
  chains         = 2,
  iter           = 2000,
  cores          = 2,
  threads        = brms::threading(4),
  control        = list(adapt_delta = 0.99)
)

tidy_model(fit_cmd, include_random = TRUE) %>% print(n = 20)
```

---

## 6. Poststratification (MrP)

Multilevel regression with poststratification (MrP) reweights model predictions to match a
population census frame, correcting for non-representative samples.

Requires a Bayesian multilevel model — set `RBESD_RUN_BAYES_VIGNETTE=true` to evaluate.

### Load the poststratification frame

`ps_demo` is a bundled census-style data frame of population cell counts for each
country × gender × age combination:

```{r}
data("ps_demo", package = "rbesd")
print(ps_demo, n = 10)
```

### Fit a Bayesian multilevel model

Context-specific predictors are not supported for MrP — use `common` predictors only:

```{r, eval = requireNamespace("brms", quietly = TRUE) && Sys.getenv("RBESD_RUN_BAYES_VIGNETTE") == "true"}
fit_mrp <- besd_regress(
  y,
  outcome        = "sp_peer",
  predictors     = list(common = c("dem_gen", "dem_age")),
  scope          = "multilevel",
  engine         = "bayes",
  random_slopes  = TRUE,
  correlated_re  = FALSE,
  backend        = "cmdstanr",
  chains         = 2,
  iter           = 1000,
  cores          = 2,
  threads        = brms::threading(4),
  control        = list(adapt_delta = 0.99)
)
```

### Build the poststrat frame and get fitted probabilities

```{r, eval = requireNamespace("brms", quietly = TRUE) && Sys.getenv("RBESD_RUN_BAYES_VIGNETTE") == "true"}
frame  <- besd_poststrat_frame(ps_demo, fit = fit_mrp, pop_col = "n_pop")
fitted <- besd_fitted_probs(fit_mrp, newdata = frame)
```

### Poststratify

Population-weighted national estimates (one row per country):

```{r, eval = requireNamespace("brms", quietly = TRUE) && Sys.getenv("RBESD_RUN_BAYES_VIGNETTE") == "true"}
mrp_national <- besd_poststratify(fitted, frame)
print(mrp_national)
```

Estimates broken down by demographic subgroup:

```{r, eval = requireNamespace("brms", quietly = TRUE) && Sys.getenv("RBESD_RUN_BAYES_VIGNETTE") == "true"}
mrp_by_dem <- besd_poststratify(fitted, frame, by = c("dem_gen", "dem_age"))
print(mrp_by_dem)
```

Add a pooled "Overall" row across all countries:

```{r, eval = requireNamespace("brms", quietly = TRUE) && Sys.getenv("RBESD_RUN_BAYES_VIGNETTE") == "true"}
mrp_overall <- besd_poststratify(fitted, frame, overall = TRUE)
print(mrp_overall)
```

Retrieve the full posterior draw distribution per group (Bayesian models only):

```{r, eval = requireNamespace("brms", quietly = TRUE) && Sys.getenv("RBESD_RUN_BAYES_VIGNETTE") == "true"}
out <- besd_poststratify(fitted, frame, post_probs = TRUE)
out$estimates   # median + 95% credible interval
out$post_probs  # list-column of posterior draws per group
```

---

## Where to go next

- Explore the BeSD Explorer Shiny app: `launch_explorer()`
- Prepare Explorer data from your own `besd_data` object: `prepare_explorer_data(dat)`
- Read `?besd_regress`, `?tidy_model`, and `?besd_poststratify` for full parameter documentation
