---
title: "rbesd workflow: dictionaries, data preparation, summaries, and regression modelling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rbesd workflow: dictionaries, data preparation, summaries, and regression modelling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
set.seed(1)
```

This vignette walks through a realistic analysis workflow in **rbesd**:

- how the **dictionary** controls harmonisation (item IDs, allowed levels, item types)
- how to convert raw exports into a validated `besd_data` object with `as_besd()`
- how to add / extend **socio-demographics** (predictors) with a dictionary
- how to create weighted summaries and plots
- how to run **by-country** and **multilevel** regressions
- how the multilevel API separates **common** vs **context** predictors
- how to fit Bayesian models with either **MCMC** or a faster **Laplace approximation**

To keep the vignette fast and robust, the Bayesian sections are shown as runnable code
but evaluated only when you opt in (see “Running Bayesian examples” below).

```{r}
library(rbesd)
library(dplyr)
```

## 1. Dictionaries: the contract for harmonisation

The BeSD dictionary is a table describing each item:

- `item_id`: the standard name used in analyses (`tf_benefits`, `sp_peer`, …)
- `item_type`: `binary`, `ordinal`, `multichoice`, …
- `levels`: allowed response labels (used for validation + ordering)
- `reverse`: whether an item should be treated as reverse-coded in “top-box” style summaries
- the question text (long and short)

```{r}
besd_dict <- besd_dictionary()
besd_dict %>% 
  select(item_id, domain, item_type, reverse) %>% 
  head(10)
```

Inspect a single item’s allowed levels:

```{r}
besd_dict %>% 
  filter(item_id %in% c("tf_benefits", "sp_peer", "pr_reasons_ease_access")) %>% 
  select(item_id, item_type, levels)
```

There is also a small built-in demographic dictionary (`dem_dictionary()`) that you can
extend with additional predictors:

```{r}
dem_dict <- dem_dictionary()
dem_dict
```

## 2. A toy “raw export” dataset included with the package

We’ll use a synthetic dataset shipped with **rbesd** via `besd_toy_raw()`. It deliberately looks like a raw export:

- raw question column names (`Q_*`)
- text responses (sometimes including common missing tokens)
- checkbox-style columns for a multichoice item
- a `weight` column and a respondent ID

```{r}
raw <- besd_toy_raw()
raw %>% glimpse()
```

## 3. Converting raw exports to `besd_data` with `as_besd()`

### 3.1 Mapping raw columns to standard `item_id`s

`as_besd()` can take a **mapping** from raw columns to standard dictionary IDs.

- names = raw column names
- values = dictionary `item_id`

```{r}
mapping <- c(
  Q_benefits     = "tf_benefits",
  Q_safety       = "tf_safety",
  Q_peer_norm    = "sp_peer",
  # socio-demographics (predictors)
  Q_gender       = "dem_gen",
  Q_age          = "dem_age",
  Q_education    = "dem_edu",
  Q_district     = "ctx_district",
  # multichoice checkboxes (multiple raw cols -> one multichoice item_id)
  Q_access_nothing       = "pr_reasons_ease_access",
  Q_access_clinic_hard   = "pr_reasons_ease_access",
  Q_access_opening_times = "pr_reasons_ease_access",
  Q_access_turned_away   = "pr_reasons_ease_access",
  Q_access_wait_long     = "pr_reasons_ease_access"
)
mapping
```

### 3.2 Extending the demographic dictionary (adding education + a context predictor)

The built-in `dem_dictionary()` contains only gender and age. Here we extend it with:

- `dem_edu`: an education predictor
- `ctx_district`: a within-country context predictor (used later in multilevel models)

We do this by appending rows with the required dictionary columns.

```{r}
dem_ext <- modify_dictionary(
  df = tibble::tibble(
    item_id = c("dem_edu", "ctx_district"),
    domain = c("predictor", "predictor"),
    item_type = c("categorical", "categorical"),
    levels = list(
      c("Primary", "Secondary", "Tertiary", "Don't know"),
      c("North", "Central", "South", "East", "West", "Capital",
        "Coastal", "Inland", "Islands", "Urban belt", "Rural belt", "Highlands")
    ),
    reverse = c(NA, NA),
    question = c("What is the highest level of education completed?",
                 "Which district do you live in?"),
    question_short = c("Education", "District")
  ),
  dict = dem_dictionary(),
  replace = FALSE
)

dem_ext
```

### 3.3 Multichoice specs: packing checkbox columns

The BeSD dictionary treats `pr_reasons_ease_access` as a **multichoice** item: in raw exports,
it is often represented as multiple checkbox columns. To convert checkboxes into a single packed
multichoice column, provide a `multichoice_specs` entry describing:

- encoding (`"binary"`: Yes/No markers)
- which label each checkbox corresponds to (`col_labels`)

To avoid typos, we fetch the canonical labels from the dictionary:

```{r}
access_levels <- besd_dict$levels[[match("pr_reasons_ease_access", besd_dict$item_id)]]
access_levels[1:5]
```

```{r}
multichoice_specs <- list(
  pr_reasons_ease_access = list(
    encoding = "binary",
    selected_values = c("yes", "selected", "true", "1"),
    unselected_values = c("no", "not selected", "false", "0"),
    col_labels = access_levels[1:5] # aligns with the 5 Q_access_* columns in mapping
  )
)
```

### 3.4 Create the `besd_data` object (with validation)

Now we run `as_besd()`:

- `missing_tokens` declares common “missing” responses
- `missing_action = "na"` turns them into `NA`
- `unknown_action = "error"` stops if something isn’t in the dictionary (useful for QA)

```{r}
besd <- as_besd(
  raw,
  mapping = mapping,
  dem_dict = dem_ext,
  multichoice_specs = multichoice_specs,
  country_col = "country",
  weight_col = "weight",
  id_col = "resp_id",
  missing_tokens = c("Don't know", "Refused"),
  missing_action = "na",
  unknown_action = "error"
)

besd
besd_validate(besd, strict = TRUE)
```

## 4. What happens when data aren’t in the right form?

### 4.1 No mapping (or no standard item columns) → error

```{r, error=TRUE}
as_besd(raw)
```

### 4.2 Unknown values → error (or coercion to `NA`)

If a value is not in the dictionary levels, `unknown_action = "error"` stops early.

```{r, error=TRUE}
raw_bad <- raw
raw_bad$Q_peer_norm[1] <- "Maybe"
as_besd(
  raw_bad,
  mapping = mapping,
  dem_dict = dem_ext,
  multichoice_specs = multichoice_specs,
  missing_tokens = c("Don't know", "Refused"),
  unknown_action = "error"
)
```

If you’re doing exploratory work, you can instead coerce unknown values to `NA`:

```{r, warning=TRUE}
besd_soft <- as_besd(
  raw_bad,
  mapping = mapping,
  dem_dict = dem_ext,
  multichoice_specs = multichoice_specs,
  missing_tokens = c("Don't know", "Refused"),
  unknown_action = "na"
)
```

### 4.3 Keeping missing tokens as explicit levels

Sometimes you want missing codes to remain visible in summaries (e.g., “Don’t know”).
Use `missing_action = "keep"`:

```{r}
besd_keep_missing <- as_besd(
  raw,
  mapping = mapping,
  dem_dict = dem_ext,
  multichoice_specs = multichoice_specs,
  missing_tokens = c("Don't know", "Refused"),
  missing_action = "keep",
  unknown_action = "error"
)

table(besd_keep_missing$sp_peer, useNA = "ifany")
```

## 5. Summaries and plots

### 5.1 Response distributions (weighted)

`summary(besd_data)` returns a `besd_summary_tbl` you can plot directly.

```{r}
sum_tbl <- summary(besd, include_demographics = TRUE)
sum_tbl %>% 
  select(country, item_id, response, pct, lcl, ucl) %>% 
  head(12)
```

Bar plots (filtering to a few items):

```{r}
plot_besd_bars(
  sum_tbl %>% filter(item_id %in% c("tf_benefits", "tf_safety", "sp_peer"))
)
```

Multichoice plots:

```{r}
plot_besd_multichoice_bars(sum_tbl, item_ids = "pr_reasons_ease_access")
```

### 5.2 Top-box summaries and spider plots

Top-box is often “Yes” for binary items or the highest agreement/importance category for ordinal items.

```{r}
tb <- besd_topbox(sum_tbl)
tb %>% select(country, item_id, topbox, pct) %>% head(10)
```

```{r}
plot_besd_spider(
  sum_tbl,
  item_ids = c("tf_benefits", "tf_safety", "sp_peer"),
  compare = "all"
)
```

## 6. Regression modelling

### 6.1 By-country regressions (fast defaults)

By-country models fit the same specification separately within each country.

```{r}
fit_bc <- besd_regress(
  besd,
  outcome = "sp_peer",
  predictors = c("dem_gen", "dem_age", "dem_edu"),
  scope = "by_country",
  engine = "frequentist"
)

tidy_bc <- besd_tidy(fit_bc, exponentiate = TRUE)
tidy_bc %>% 
  filter(effect_type == "fixed") %>% 
  arrange(country, term) %>% 
  head(12)
```

### 6.2 Multilevel regressions: common vs context predictors

In multilevel mode, you can supply predictors as a list with two roles:

- `common`: effects that are assumed to be *comparable across countries*  
  (and can optionally have random slopes by country)
- `context`: within-country predictors whose interpretation can be country-specific  
  (rbesd expands these into country-specific contrasts, with within-country reference levels)

```{r}
preds <- list(
  common = c("dem_gen", "dem_age", "dem_edu"),
  context = c("ctx_district")
)
```

A multilevel frequentist logistic regression requires **lme4**. If it’s installed:

```{r, eval = requireNamespace("lme4", quietly = TRUE)}
fit_ml <- besd_regress(
  besd,
  outcome = "sp_peer",
  predictors = preds,
  scope = "multilevel",
  engine = "frequentist",
  random_slopes = TRUE
)

tidy_ml <- besd_tidy(fit_ml, exponentiate = TRUE)
tidy_ml %>% 
  filter(effect_type == "fixed") %>% 
  head(12)
```

To see how context predictors were expanded, inspect the term map:

```{r}
# The term map describes how model terms relate back to variables, levels, and countries.
# (This is useful for reporting and for post-processing coefficients.)
if (exists("fit_ml")) {
  nm <- names(fit_ml$meta$term_map)
  head(nm[grepl("^ctx_", nm)], 10)
}
```

### 6.3 Multichoice outcomes in regression

If you regress a multichoice outcome, `besd_regress()` automatically expands it into
multiple binary outcomes (one per option), and fits each one.

```{r, eval = FALSE}
fit_mc <- besd_regress(
  besd,
  outcome = "pr_reasons_ease_access",
  predictors = c("dem_gen", "dem_age", "dem_edu"),
  scope = "by_country",
  engine = "frequentist"
)

# `fit_mc$meta$outcomes` contains the expanded outcome names
fit_mc$meta$outcomes
```

## 7. Bayesian modelling: MCMC and Laplace

Bayesian regression uses **brms**. rbesd forwards `...` to `brms::brm()`, so you can control:

- backend (`backend = "cmdstanr"` recommended)
- algorithm (`"sampling"` for MCMC; `"laplace"` for fast approximation)
- priors, chains, iter, adapt_delta, etc.

### Running Bayesian examples

To evaluate the Bayesian chunks in this vignette, set:

- `RBESD_RUN_BAYES_VIGNETTE=true`

### 7.1 MCMC sampling (HMC)

```{r, eval = requireNamespace("brms", quietly = TRUE) && Sys.getenv("RBESD_RUN_BAYES_VIGNETTE") == "true"}
fit_bayes_mcmc <- besd_regress(
  besd,
  outcome = "sp_peer",
  predictors = preds,
  scope = "multilevel",
  engine = "bayes",
  random_slopes = TRUE,
  correlated_re = FALSE,   # use `||` in the random-effects term (often faster)
  backend = "cmdstanr",
  chains = 2,
  iter = 1000,
  seed = 1,
  refresh = 0
)

besd_tidy(fit_bayes_mcmc, exponentiate = TRUE) %>% 
  filter(effect_type == "fixed") %>% 
  head(12)
```

### 7.2 Laplace approximation (fast, approximate)

Laplace can be dramatically faster than full MCMC for larger datasets, at the cost of
being an approximation. It is most commonly used via the **cmdstanr** backend.

```{r, eval = requireNamespace("brms", quietly = TRUE) && Sys.getenv("RBESD_RUN_BAYES_VIGNETTE") == "true"}
fit_bayes_laplace <- besd_regress(
  besd,
  outcome = "sp_peer",
  predictors = preds,
  scope = "multilevel",
  engine = "bayes",
  random_slopes = TRUE,
  correlated_re = FALSE,
  backend = "cmdstanr",
  algorithm = "laplace",
  seed = 1,
  refresh = 0
)

besd_tidy(fit_bayes_laplace, exponentiate = TRUE) %>% 
  filter(effect_type == "fixed") %>% 
  head(12)
```

## 8. Where to go next

Typical next steps for real projects:

- tailor dictionaries with `modify_dictionary()` (e.g. translated labels; survey-specific options)
- add more predictors via a demographic/predictor dictionary and pass them to `as_besd()`
- use multilevel models with:
  - richer `common` predictors (individual-level factors)
  - `context` predictors describing within-country strata (district, facility type, sampling frame)
- for Bayesian fits:
  - start with `algorithm = "laplace"` to iterate quickly
  - switch to full MCMC (`algorithm = "sampling"`) for final inference and diagnostics
