---
title: "BeSD Country Report: `r params$country`"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    self_contained: true
params:
  country: "Unknown"
  besd_sum: NULL
  demo_sum: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 7)
library(rbesd)
library(dplyr)
library(ggplot2)

cty      <- params$country
besd_sum <- params$besd_sum
demo_sum <- params$demo_sum
```

# `r params$country`

This report summarises BeSD (Behavioural and Social Drivers of vaccination)
survey findings for **`r params$country`**.

---

## Response Distributions

The chart below shows the distribution of responses to each BeSD indicator item.

```{r bar_plot, fig.height=max(6, dplyr::n_distinct(dplyr::filter(besd_sum, country == cty)$item_id) * 0.6)}
dd <- besd_sum |> dplyr::filter(country == cty)
if (nrow(dd) > 0) {
  rbesd::plot_besd_bars(dd)
} else {
  cat("No response data available for", cty)
}
```

---

## Top-Box Summary (Radar)

```{r radar_data}
tb <- rbesd::besd_topbox(besd_sum |> dplyr::filter(country == cty))
if (nrow(tb) > 0) {
  tb |>
    dplyr::arrange(domain, item_id) |>
    dplyr::select(Domain = domain, Item = item_id,
                  Question = question_short, `Top-box %` = pct) |>
    knitr::kable(digits = 1)
} else {
  cat("No top-box data available for", cty)
}
```

---

## Demographic Profile

```{r demo_plot, fig.height=5}
tryCatch(
  rbesd::plot_besd_demographics_table(demo_sum, country = cty),
  error = function(e) {
    message("No demographic data available for ", cty)
    invisible(NULL)
  }
)
```

---

*Report generated by the BeSD Explorer &middot; IMMS &middot; `r format(Sys.time(), "%Y-%m-%d %H:%M")`*
